---
output:
  html_document: default
  pdf_document: default
---
RED WINE EXPLORATION by PIERRE CONREAUX
========================================================

# Introduction

Our goal is to get a better understanding of which chemical properties influence ***the quality of red wines.***

We will use a public dataset available which come from:
*P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis.* 
*Modeling wine preferences by data mining from physicochemical properties.*
*In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.*

For more information visit the following link:
https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt

First, we load ***the necessary packages***:

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2) # plotting system
library(knitr) # report generation
library(dplyr) # grammar data manipulation
library(corrplot) # plot correlation
```

Then, we ***load the data:***

```{r echo=FALSE, Load_the_Data}
wine<-read.csv('wineQualityReds.csv',check.names = FALSE, row.names = 1)
``` 

To confirm the loading, we ***read the head*** of the data:

```{r echo=FALSE, head}
head(wine)
```

# Univariate Plots Section

Before plotting, we need to understand our data sets and its variables.

After a quick description and a summary, we will plot each variables.

## Data set

```{r echo=FALSE, dimension}
dim(wine)
```

This tidy data set contains 1,599 red wines with 11 variables on the chemical properties of the wine, and 1 variable on the quality of the wine. 

Regarding the quality variable, at least 3 wine experts rated the quality of each wine, providing a rating between 0 (very bad) and 10 (very excellent).

Noticed that our data is already tidy, so we do not have to clean it.


## Variables

```{r echo=FALSE, columns}
names(wine)
```


### Description of variables:

   1 - fixed acidity: most acids involved with wine or fixed or nonvolatile (do not evaporate readily)

   2 - volatile acidity: the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste

   3 - citric acid: found in small quantities, citric acid can add 'freshness' and flavor to wines

   4 - residual sugar: the amount of sugar remaining after fermentation stops, it's rare to find wines with less than 1 gram/liter and wines with greater than 45 grams/liter are considered sweet

   5 - chlorides: the amount of salt in the wine

   6 - free sulfur dioxide: the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine

   7 - total sulfur dioxide: amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

   8 - density: the density of water is close to that of water depending on the percent alcohol and sugar content

   9 - pH: describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale

   10 - sulphates: a wine additive which can contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and antioxidant

   11 - alcohol: the percent alcohol content of the wine

   Output variable (based on sensory data): 
   12 - quality (score between 0 and 10)


### data summary

```{r  echo=FALSE, structure }
str(wine)
```


```{r echo=FALSE, summary}
summary.data.frame(wine)
```


## Plots

We start by create a ***function to create histogram***:

```{r echo=FALSE, histogram_function}
histogram <- function(variable,...){
  ggplot(data = wine, aes_q(as.name(variable)))+
    geom_histogram(...)
}
```

### Fixed Acidity

```{r echo=FALSE, Histogram_Fixed_Acidity}
histogram("fixed.acidity", binwidth = 0.2)+
  scale_x_continuous(breaks = 4:16)
```

The graph depicts ***the distribution of fixed acidity.***

We can see that the distribution is ***right skewed.***

```{r echo=FALSE, summary_fixed.acidity}
summary(wine$fixed.acidity)
```


### Volatile Acidity

```{r echo=FALSE, Histogram_Volatile_Acidity}
histogram("volatile.acidity", binwidth = 0.02)+
  scale_x_continuous(breaks = seq(0, 1.6, 0.1))
```

The graph depicts ***the distribution of volatile acidity.***

We can see that the distribution is approximately ***normally skewed.***

```{r echo=FALSE, summary_volatile.acidity}
summary(wine$volatile.acidity)
```


### Citric acid

```{r echo=FALSE, Histogram_Citric_Acid}
histogram("citric.acid", binwidth = 0.02)+
  scale_x_continuous(breaks = seq(0, 1.6, 0.1))
```

The graph depicts ***the distribution of citric acid.***

We can see that the distribution is ***multi-modal.***

```{r echo=FALSE, summary_citric.acid}
summary(wine$citric.acid)
```


### Residual Sugar

```{r echo=FALSE, Histogram_Residualc_Sugar}
histogram("residual.sugar", binwidth = 0.2)+
  scale_x_continuous(breaks = 0:16)+
  coord_cartesian(xlim = c(1, 7)) #lim to keep visibility
```

The graph depicts ***the distribution of residual sugar.***

We can see that the distribution is approximately ***right skewed.***

```{r echo=FALSE, summary_residual.sugar}
summary(wine$residual.sugar)
```


### Chlorides

```{r echo=FALSE, Histogram_Chlorides}
histogram("chlorides", binwidth = 0.01)+
  scale_x_continuous(breaks = seq(0, 0.7, 0.1))+
  coord_cartesian(xlim = c(0, 0.3)) #lim to keep visibility
```

The graph depicts ***the distribution of chlorides.***

We can see that the distribution is  ***normal.***

```{r echo=FALSE, summary_chlorides}
summary(wine$chlorides)
```


### Free sulfur dioxide

```{r echo=FALSE, Histogram_Free_Sulfur_Dioxide}
histogram("free.sulfur.dioxide", binwidth = 1)+
  scale_x_continuous(breaks = seq(0, 72, 5))
```

The graph depicts ***the distribution of free sulfur dioxide.***

We can see that the distribution is  ***right skewed.***

```{r echo=FALSE, summary_free.sufur.dioxide}
summary(wine$free.sulfur.dioxide)
```


### Total sulfur dioxide

```{r echo=FALSE, Histogram_Total_Sulfur_Dioxide}
histogram("total.sulfur.dioxide", binwidth = 3)+
  scale_x_continuous(breaks = seq(0, 289, 20))+
  coord_cartesian(xlim = c(6, 170)) #lim to keep visibility
```

The graph depicts ***the distribution of total sulfur dioxide.***

We can see that the distribution is  ***right skewed.***

```{r echo=FALSE, summary_total.sulfur.dioxide}
summary(wine$total.sulfur.dioxide)
```


### Density


```{r echo=FALSE, Histogram_Density}
histogram("density", binwidth = 0.0005)+
  scale_x_continuous(breaks = seq(0.990, 1.005, 0.0025))
```

The graph depicts ***the distribution of density.***

We can see that the distribution is  ***normal.***

```{r echo=FALSE, summary_density}
summary(wine$density)
```


### pH

```{r echo=FALSE, Histogram_pH}
histogram("pH", binwidth = 0.02)+
  scale_x_continuous(breaks = seq(0, 4.070, 0.1))
```

The graph depicts ***the distribution of pH.***

We can see that the distribution is  ***normal.***

```{r echo=FALSE, summary_pH}
summary(wine$pH)
```


### Sulphates

```{r echo=FALSE, Histogram_Sulphates}
histogram("sulphates", binwidth = 0.02)+
  scale_x_continuous(breaks = seq(0, 2, 0.1))+
  coord_cartesian(xlim = c(0.3, 1.4)) #lim to keep visibility
```

The graph depicts ***the distribution of sulphates.***

We can see that the distribution is  ***right skewed.***

```{r echo=FALSE, summary_sulphates}
summary(wine$sulphates)
```


### Alcohol

```{r echo=FALSE, Histogram_Alcohol}
histogram("alcohol", binwidth = 0.2)+
  scale_x_continuous(breaks = seq(0, 15, 1))+
  coord_cartesian(xlim = c(8.4, 14.9)) #lim to keep visibility
```

The graph depicts ***the distribution of alcohol.***

We can see that the distribution is  ***right skewed.***

```{r echo=FALSE, summary_alcohol}
summary(wine$alcohol)
```


### Quality

```{r echo=FALSE, Histogram_Quality}
histogram("quality", binwidth = 1)
  scale_x_continuous(breaks = seq(0, 8, 1))
```

The graph depicts ***the distribution of quality.***

We can see that the distribution is  ***normal.***

```{r echo=FALSE, summary_quality}
summary(wine$quality)
```


### Rating

To ensure a better visibility in multivariate section, we create ***rating, a discrete variable***, to gather quality by levels.  

```{r echo=FALSE, rating}
wine$rating <- cut(wine$quality, breaks = c(3, 4, 6, 8), 
                          include.lowest = T, right = T, 
                          labels = c("Poor", "Medium", "Excellent"))
```


To plot a discrete variable, as rating, we need to modify the histogram code.

```{r echo=FALSE, Histogram_rating}

ggplot(data = wine, aes(x = factor(rating))) +
  geom_bar() +
  xlab('rating')
```

```{r echo=FALSE, summary_rating}
summary(wine$rating)
```

# Univariate Analysis


### What is the structure of your dataset?

This data set contains ***1,599 red wines with 13 columns.***

***11 numeric variables*** on the chemical properties of the wine and ***2 quality variable.***


### What is/are the main feature(s) of interest in your dataset?

The main feature is the ***quality*** of the wine.


### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Each numeric variables represent a charateristic of the wine, so ***each variables have to be investigating***.


### Did you create any new variables from existing variables in the dataset?

We created a rating variable that we will use in multivariate plots section.


### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

The data set is already tidy, so we not have to perform any operations to tidy it.


# Bivariate Plots Section

In this section, we will start by analysis the relation between quality and the other variables.

After, we will study interaction between features.


## quality 

In order to be able to create boxplots we need to make ***a discrete variable for quality.***

We will do this in using ***cut:***

```{r echo=FALSE, discrete_quality}
quality.d <- cut(wine$quality, seq(0,10,1), labels=c(1:10))
```

Now we can create a ***function to make boxplots.***

```{r echo=FALSE, function_box_plot}
boxplot <- function(variable,...){
  ggplot(aes_q(x = ~quality.d, y = as.name(variable)), data = wine) +
  geom_boxplot()+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)
  
}
```

Finally, we can ***combine our functions.***


### Quality vs fixed.acidity 

```{r echo=FALSE, quality_vs_fixed_acidity}

boxplot('fixed.acidity')
```

The median level of fixed acidity seems close between the quality rating.
We can notice a sligh rise between 6 and 7.

```{r echo=FALSE, correlation_quality_fixed.acidity}
cor(wine$quality,wine$fixed.acidity)
```


### Quality vs volatile.acidity 


```{r echo=FALSE, quality_vs_volatile_acidity}

boxplot('volatile.acidity')
```

The volatile acidity is negatively correlated to the quality.

We can see a stability of the level of volatile acidity between 7 and 8, 
nevertheless this is not significant, because a small number of wines
have a rating quality of 8.

```{r echo=FALSE, correlation_quality_volatile.acidity}
cor(wine$quality,wine$volatile.acidity)
```


### Quality vs citric.acid

```{r echo=FALSE, quality_vs_citric.acid}

boxplot('citric.acid')
```

The citric acid is positively correlated with quality.

```{r echo=FALSE, correlation_quality_citric.acid}
cor(wine$quality,wine$citric.acid)
```


### Quality vs residual sugar

```{r echo=FALSE, quality_vs_residual.sugar}
boxplot('residual.sugar')
```

The residul sugar seems weakly correalated to quality.

```{r echo=FALSE, correlation_quality_residual.sugar}
cor(wine$quality,wine$residual.sugar)
```


### Quality vs chlorides

```{r echo=FALSE, quality_vs_chlorides}
boxplot('chlorides')
```

The chlorides seem weakly correalated to quality.

```{r echo=FALSE, correlation_quality_chlorides}
cor(wine$quality,wine$chlorides)
```


### Quality vs free.sulfur.dioxide

```{r echo=FALSE, quality_vs_free.sulfur.dioxide}

boxplot('free.sulfur.dioxide')
```

Free sulfur dioxide and quality seem positively correlated between 3 and 5,
then negatively correlated.

```{r echo=FALSE, correlation_quality_free.sulfur.dioxide}
cor(wine$quality,wine$free.sulfur.dioxide)
```


### Quality vs total.sulfur.dioxide

```{r echo=FALSE, quality_vs_total.sulfur.dioxide}
boxplot('total.sulfur.dioxide')
```

Total sulfur dioxide and quality seem positively correlated between 3 and 5,
then negatively correlated.

```{r echo=FALSE, correlation_quality_total.sulfur.dioxide}
cor(wine$quality,wine$total.sulfur.dioxide)
```


### Quality vs density

```{r echo=FALSE, quality_vs_density}
boxplot('density')
```

Density seems to be negatively correlated with quality.

Nevertheless, we can notice a slight increase between 4 and 5.

```{r echo=FALSE, correlation_quality_density}
cor(wine$quality,wine$density)
```


### Quality vs pH

```{r echo=FALSE, quality_vs_pH}
boxplot('pH')
```

PH seems to be negatively correlated with quality.

```{r echo=FALSE, correlation_quality_pH}
cor(wine$quality,wine$pH)
```


### Quality vs sulphates

```{r echo=FALSE, quality_vs_sulphates}
boxplot('sulphates')
```

Sulphates and quality seems to be positevely correlated.

```{r echo=FALSE, correlation_quality_sulphates}
cor(wine$quality,wine$sulphates)
```


### Quality vs alcohol

```{r echo=FALSE, quality_vs_alcohol}
boxplot('alcohol')
```

Alcohool and quality seems to be positevely correlated, except for the rate 5 which have a lot of ouliers.

```{r echo=FALSE, correlation_quality_fixed.alcohol}
cor(wine$quality,wine$alcohol)
```


## Complementary bivariate plots

In order to choose some bivariate plots, we will start to create a correlation matrix, then visualize it.

```{r echo=FALSE, correlation_matrix}
res<-cor(wine[0:12], method = c("spearman"))
```


```{r echo=FALSE, visualize_correlation}
corrplot(res, type="upper",tl.col="black")
```

We will plot the following bivariate correlations to get a better understanding:`
- fixed acidity and citric acid 
- fixed acidity and density
- fixed acidity and ph
- free sulfur dioxide and total sulfur dioxide


***Function to create scatterplot:*** 

```{r echo=FALSE, scatterplot_function}
scatter_plot<- function(x,y){
  ggplot(aes_q(x = as.name(x),y = as.name(y)),
         data=wine)+
    geom_point(alpha = 0.3, size = 1)+
    stat_smooth(method='lm')
}
```


### Fixed acidity vs citric acid


```{r echo=FALSE,scatter_fixed.acidity_citric.acid}
scatter_plot(x = "fixed.acidity",y="citric.acid")
```

The scatterplot shows a linear positive moderate strong relation, nevertheless we can observe many outliers.

```{r echo=FALSE, correlation_citric.acid_fixed.acidity}
cor(wine$citric.acid,wine$fixed.acidity)
```


### Fixed acidity and density

```{r echo=FALSE, scatter_fixed.acidity_density}
scatter_plot(x = "fixed.acidity",y="density")
```

The scatterplot shows a linear positive strong relation.

```{r echo=FALSE, correlation_density_fixed.acidity}
cor(wine$density,wine$fixed.acidity)
```


### Fixed acidity vs pH

```{r echp=FALSE, scatter_fixed/acidity_pH}
scatter_plot(x = "fixed.acidity",y="pH")
```

The scatterplot shows a linear negative strong relation.

```{r echo=FALSE, correlation_pH_fixed.acidity}
cor(wine$pH,wine$fixed.acidity)
```


### Free sulfur dioxide vs total sulfur dioxide

```{r echo=FALSE, scatter_free_and_total}
scatter_plot(x="free.sulfur.dioxide",y="total.sulfur.dioxide")
```

The scatterplot shows a linear positive strong relation.

```{r echo=FALSE, correlation_total.sulfur.dioxide_free.sulfur.dioxide}
cor(wine$total.sulfur.dioxide,wine$free.sulfur.dioxide)
```


# Bivariate Analysis

Correlation summary:

```{r echo=FALSE, visualize_correlation_summary}
corrplot(res, type="upper",tl.col="black")
```

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Quality is positevely correlated with alcohol and sulpates and negatively with volatile acidity. 

The other relations are not sufficiently correlated.


### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

We observed the following relations:
- a positive correlation between acid citric and fixed acidity.
- a positve correlation between density and fixed acidity.
- a negative correlation between pH and fixed acidity.
- a stong positive relation between free sulfur dioxide and total sulfur dioxide.


### What was the strongest relationship you found?

The strongest relationship is between free sulfur dioxide and total sulfur dioxide,
which is not really surprising, considering that total sulfur dioxide contains free sulfur dioxide.


# Multivariate Plots Section


***Funtion to create multivariate function:***

```{r echo=FALSE, function_multivariate}
multivariate <- function(x, y){
  ggplot(wine, aes_q(x = as.name(x), y = as.name(y), color = ~ rating))+
    geom_point(alpha = 1, size = 1)+
    scale_color_brewer(type = "seq", palette = "YlOrRd")+
        geom_smooth(method = "lm")

  }
```


## Alcohol, volatile acidity and quality

```{r echo=FALSE, alcohol_volatile.acidity_quality}
multivariate(x="alcohol", y="volatile.acidity")
```

We can observe that excellent wines tend to have lower volatile acidity over the entire range of alcohol.


## Alcohol, volatile acidity and quality

```{r echo=FALSE, alcohol_sulphate_quality}
multivariate("alcohol", "sulphates")
```

We can observe that excellent wines tend to have higher sulphates over the entire range of alcohol.


## Sulphates, volatile acidiy, quality

```{r echo=FALSE, sulphate_volatile.acidity_quality}
multivariate("sulphates", "volatile.acidity")
```

We can observe that excellent wines tend to have lower volatile acidity over the entire range of sulphates.


## Fixed acidity, citric acid, quality

```{r echo=FALSE, fixed.acidity_citric.acid_quality}
multivariate("fixed.acidity", "citric.acid")
```

We can observe that excellent wines tend to have higher citic acid over the entire range of fixed acidity.


## Fixed acidity, density, quality

```{r echo=FALSE, fixed.acidity_density_quality}
multivariate("fixed.acidity", "density")
```

We can observe that excellent wines tend to have lower density over the entire range of fixed acidity.


## Fixed acidity, pH, quality

```{r echo=FALSE, fixed.acidity_pH_quality}
multivariate("fixed.acidity", "pH")
```

We do not see a clear pattern between the three variables.


## Volatile acidity, density and quality

```{r echo=FALSE, volatile.acidity_density_quality}
multivariate("volatile.acidity", "density")
```

We can observe that excellent wines tend to have lower density over the entire range of volatile acidity.

We can also see that excellent wines are represented on a lower range of volatile acidity than medium and poor.


## Free sulfur dioxide, total sulfur dioxide, quality

```{r echo=FALSE, free.sulfur.dioxide_total.sulfur.dioxide_quality}
multivariate("free.sulfur.dioxide", "total.sulfur.dioxide")
```

We do not see a clear pattern between the three variables.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

We principally explored the relationships which have the highest correlation with quality.

We observed that higher alcohol and lower volatile acidity means generally quality wines, 
that excellent wines tend to have higher citic acid over the entire range of fixed acidity.

We also observed that excellent red wines tend to have higher citic acid over the entire range of fixed acidity,
and lower density over the entire range of volatile acidity.


### Were there any interesting or surprising interactions between features?

Low density and low volatile acidity contain the better red wines.

------

# Final Plots and Summary


### Plot One
```{r echo=FALSE, Plot_One}
histogram("citric.acid", binwidth = 0.02)+
  scale_x_continuous(breaks = seq(0, 1.6, 0.1))+
  xlab('Citric Acid (g / dm^3)')+
  ylab('Red Wines')+
  ggtitle('Distribution of acid citric in Red Wines')
```


### Description One

The graph depicts the distribution of acid citric in Red Wines.

This distribution wakes up our interest, because is the only muli-modal distribution among the variables.

Thus, it will be interesting to explore is relation with the quality variable.


### Plot Two
```{r echo=FALSE, Plot_Two}
boxplot('citric.acid')+
  xlab('Citric Acid (g / dm^3)')+
  ylab('Quality of red wines')+
  ggtitle('Citric Acid is correlated positevely with red wines quality')
```


### Description Two

The box plot describes the relation between citric acid and quality of red wines.

Citric acid and qualiry of quaity of red wines have a postive correlation.

We can calcul the correlation to deepen our understanding of the two variables:

```{r echo=FALSE, correlation_citric.acid_quality}
cor(wine$quality,wine$citric.acid,method="spearman")
```

The correlation is positive but not really strong.

But it can be interesting to see the relation between citric acid and other variables.

### Plot Three
```{r echo=FALSE, Plot_Three}
multivariate("fixed.acidity", "citric.acid")+
  xlab("Fixed Acidity (tartaric acid - g / dm^3)")+
  ylab("Citric Acid (g / dm^3)")+
  labs(colour="Quality of Red Wines")+
  ggtitle("Excellent wines contain more citric acid by fixed acidity that poor wines.")
```


### Description Three

The multivariate graph represents the relation between citric acid, fixed acidity and quality.

We can observe that for each level of fixed acidity, excellent wines contain more citric acid than the two other categories (medium and poor).

Excellent wines conatin more citric acid by fixed acidity that poor wines.


------

# Reflection

Our project was to explore the variables composing red wines, and particullary their impacts on the quality of red wine.

In a first time, we plotted each distributions variables. 
The vast majority of variable were right skewed, but we found out a mutivariate disbrution, citric acid, which attrated our curiosity.
We also found out a limitation at our analysis: the distribution of the wines quality is largely composed by two rating: 5 and 6

In a second time, in order to not miss something, we plotted each variables with the quality variable. 
We discover the following cues:
- a positive correlation with acid citric, alcohol and sulphates.
- a negative correaltion with volatile acidity.

In parallel we observed the following relations:
- a positive correlation between acid citric and fixed acidity.
- a positve correlation between density and fixed acidity.
- a negative correlation between pH and fixed acidity.
- a stong positive relation between free sulfur dioxide and total sulfur dioxide.

At this step it was difficult to analyze the factor to a good wines, because the correlation between the variables and the quality was not strong enough.

In order to get a better understanding, we chosen a number of multivariate to plot.
The most notables were:
- Alcohol, volatile acidity and quality
- Alcohol, volatile acidity and quality
- Sulphates, volatile acidiy, quality
- Fixed acidity, density, quality
- Volatile acidity, density and quality

And of course Fixed acidity, citric acid and quality which ended our reflection on citric acid and quality.

We observed that for each level of fixed acidity, excellent wines containned more citric acid than the two other categories of red wines (medium and poor).

To improve our analysis we can collect more data, especially wines with a rating different of 5 and 6.

# Ressources

Info on the data set: https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt
Plot: https://docs.google.com/document/d/1-f3wM3mJSkoWxDmPjsyRnWvNgM57YUPloucOIl07l4c/pub
Histogram: http://www.sthda.com/english/wiki/ggplot2-histogram-plot-quick-start-guide-r-software-and-data-visualization#basic-histogram-plots
Correlation: http://www.sthda.com/english/wiki/correlation-matrix-a-quick-start-guide-to-analyze-format-and-visualize-a-correlation-matrix-using-r-software
Multivariate analysis: http://little-book-of-r-for-multivariate-analysis.readthedocs.io/en/latest/src/multivariateanalysis.html





